---
title: "Differential expression and gene set enrichment analysis"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center', fig.height = 5)
```

# Setup

```{r warning=FALSE, message=FALSE}
library(readxl)
library(stringr)
library(data.table)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(ggplot2)
library(ggbiplot)
library(ggrepel)
library(ggridges)
library(ggpubr)
library(gridExtra)
library(genefilter)
library(enrichplot)
library(ggraph)
library(igraph)
library(tidygraph)
library(pathview)
library(gage)
library(AnnotationDbi)
library(biomaRt)
```

```{r}
theme_set(theme_pubr(border=TRUE))
```

# Read data

```{r}
input <- file.path("data","GF vs. SPF Gene Expression Analysis.xlsx")
data <- read_excel(input, sheet = 1)
setDT(data)
head(data)
```

Remove genes w/o counts (also tried removing genes with a few counts, e.g. 5, the results are not different from this).

```{r}
count_cols <- grep("count",names(data),value=TRUE)
filtrows <- data[,rowSums(.SD),.SDcols=count_cols]==0
data <- data[!filtrows]
```

Get raw counts in `DESeqDataSet`.

```{r}
count_cols <- grep("count",names(data),value=TRUE)
count_data <- as.data.frame(data[,..count_cols])
rownames(count_data) <- data$Identifier
colnames(count_data) <- stringr::str_extract(count_cols,"(GF|SPF)\\d")
col_data <- data.frame(
  strain=factor(c(rep("GF",3),rep("SPF",3)), levels=c("SPF","GF")),
  sample=factor(colnames(count_data),levels=c("SPF1","SPF2","SPF3","GF1","GF2","GF3"))
)
dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData = col_data,
  design = ~ strain
)
dds
```

# Exploratory data analysis

## Samples comparison

```{r}
sampleDists <- dist(t(assay(dds)))
sampleDistMatrix <- as.matrix(sampleDists)
colors <- colorRampPalette(rev(brewer.pal(9,"Blues")))(255)
phs <- pheatmap(
  sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  col=colors, 
  main = "Samples distance heatmap\n(based on raw counts)"
)
```

## Variance normalization

```{r include=FALSE}
vsd <- vst(assay(dds))
pcavsd <- prcomp(t(vsd), scale=TRUE)
df <- data.frame(t(vsd))
df$col <- c(rep("GF",3),rep("SPF",3))
df$sample <- rownames(df)
pcas <- data.table(scale(pcavsd$x[,1:2]), keep.rownames="sample")
pcav <- ggbiplot::ggbiplot(pcavsd, groups=factor(df$col), var.axes=FALSE) +
  geom_label_repel(data=pcas, mapping=aes(x=PC1, y=PC2, label=sample)) + 
  guides(colour=FALSE)
pcavscree <- ggbiplot::ggscreeplot(pcavsd) + labs(title="", x="PC")
grid.arrange(
  pcav, pcavscree, 
  ncol=2, widths=2:1, top="Samples PCA following variance normalization (VST)"
)
```

```{r}
rds <- rlog(assay(dds))
pcards <- prcomp(t(rds), scale=TRUE)
df <- data.frame(t(rds))
df$col <- c(rep("GF",3),rep("SPF",3))
df$sample <- rownames(df)
pcas <- data.table(scale(pcards$x[,1:2]), keep.rownames="sample")
pcar <- ggbiplot::ggbiplot(pcards, groups=factor(df$col), var.axes=FALSE) +
  geom_label_repel(data=pcas, mapping=aes(x=PC1, y=PC2, label=sample)) + 
  guides(colour=FALSE)
pcarscree <- ggbiplot::ggscreeplot(pcards) + labs(title="", x="PC")
grid.arrange(
  pcar, pcarscree, 
  ncol=2, widths=2:1, top="Samples PCA following variance normalization (rlog)"
)
```

## Clustering

Clustering by genes with most variable counts after normalization.  

First, mapping entrez ids to gene names.

```{r}
require(biomaRt)
mart <- useDataset("mmusculus_gene_ensembl", useMart("ensembl"))
genes <- getBM(
  attributes=c("ensembl_gene_id","external_gene_name","description"),
  filters="ensembl_gene_id",
  values=rownames(dds),
  mart=mart
)
```
```{r}
setDT(genes)
genenames <- cbind.data.frame(ensembl_gene_id=rownames(dds))
setDT(genenames)[genes,on='ensembl_gene_id',external_gene_name:=i.description]
```

Top 100 most variable genes on a heatmap.  

```{r fig.height=8, fig.width=4, fig.keep='all'}
require(genefilter)
heatmap_function <- function(rds,include=1:ncol(rds)) {
  topVarGenes <- head(order(rowVars(rds[,include]), decreasing = TRUE), 100)
  mat <- rds[topVarGenes,include]
  mat <- mat - rowMeans(mat)
  rownames(mat) <- stringr::str_extract(genenames$external_gene_name, "[^,\\[\\(]+")[topVarGenes]
  phg <- pheatmap(mat,treeheight_col=0,treeheight_row=0,show_rownames=FALSE)
}

heatmap_function(rds)
#heatmap_function(rds,1:5)
#heatmap_function(rds,c(1,2,4,5))
```


# DESeq2

Without SPF3.

```{r}
excols <- match(c("SPF3"),colnames(dds))
dds <- dds[,-excols]
dds <- DESeq(dds)
res <- results(dds)
```

## Compare DESeq2 and input LFC  

```{r fig.height=8}
all.equal(data$Identifier, rownames(res)) # sanity check
compadt <- data.table(
  gene=data$Name,
  input_lfc=data$`GF vs. SPF           Log fold change`,
  input_pval=data$`GF vs. SPF         P-value`,
  input_fdr=data$`GF vs. SPF FDR p-value`,
  DESeq2_lfc=res$log2FoldChange,
  DESeq2_lfcSE=res$lfcSE,
  DESeq2_pval=res$pvalue, 
  DESeq2_padj=res$padj
)
outliers <- compadt[abs(input_lfc)>10 | abs(DESeq2_lfc)>10]
ggplot(compadt[!(is.na(input_lfc)|is.na(DESeq2_lfc)|is.na(DESeq2_lfcSE))], aes(input_lfc, DESeq2_lfc)) + 
  geom_point(aes(colour=DESeq2_lfcSE), alpha=0.5) + 
  geom_abline(slope=1,intercept=0) +
  geom_label_repel(data=outliers, aes(label=gene))
data[Name %in% outliers$gene]
```

## MA plots

```{r}
plotMA(res, ylim = c(-5, 5), main="MA plot (w/o LFC shrinkage)")
```

After LFC shrinkage:

```{r}
# resultsNames(dds) # see coeffs
res <- lfcShrink(dds, coef="strain_GF_vs_SPF", type = "normal")
plotMA(res, ylim = c(-2, 2), main="MA plot (after LFC shrinkage)")
```

```{r include=FALSE, eval=FALSE}
hist(
  res$pvalue[res$baseMean > 1], 
  breaks=0:20/20, col="grey50", border="white",
  main = "Distribution of p values", x="p value"
)
```

# GAGE

Using DESeq2 DEGs.

```{r}
resdata <- as.data.table(res)
#sum(abs(resdata$log2FoldChange)>2&resdata$pvalue<0.01, na.rm=TRUE) # DE genes
exp_data <- res$log2FoldChange
names(exp_data) <- data$Name
```

Need to retrieve gene sets data from KEGG.  

```{r message=FALSE, cache=TRUE}
kegg.mm <- kegg.gsets(species="mmu")
```

Next, map genes' ENTREZ IDs to symbols.  

```{r message=FALSE}
kegg.mm.id <- lapply(kegg.mm$kg.sets, function(x) eg2id(x, category="SYMBOL", org="mouse")[,2])
all.equal(lapply(kegg.mm$kg.sets,length),lapply(kegg.mm.id,length)) # sanity check
```

### Run GAGE  

```{r}
kegg_gage <- gage(
  exp_data, 
  gsets=kegg.mm.id,
  same.dir=FALSE, 
  ref=NULL, samp=NULL
)
```

It is recommanded to set `same.dir=FALSE` for KEGG pathways analysis so as to capture gene sets i.e. pathways perturbed towards both directions simultaneously. These are pathways that are more perturbed than expected (hence the label `greater`).

```{r}
dim(kegg_gage$greater)
```

There are `r dim(kegg_gage$greater)[1]` of them.  
Next, sort and count **signficant** gene sets:  

```{r message=FALSE}
kegg_sig <- sigGeneSet(kegg_gage)
dim(kegg_sig$greater)
```

There are `r dim(kegg_sig$greater)[1]` of those.  
Extract a **non-redundant significant** gene set list:  

```{r}
nrlist <- sapply(kegg_sig[1], function(x){
  if (nrow(x)>1)
    esset.grp(x, exp_data, kegg.mm.id, same.dir=FALSE)
  else if (nrow(x)==1)
    x
  else NULL
}, simplify=FALSE, USE.NAMES=TRUE)

length(nrlist$greater$allSets) # signficant gene sets
length(nrlist$greater$essentialSets) # non-redundant signficant gene set
nrlist$greater$essentialSets
```

### Inspect overlap between genes in gene sets 

```{r}
overlaps <- nrlist$greater$overlapCounts
labels <- str_extract(nrlist$greater$allSets, "(?<=mmu\\d{5} ).*")
sizes <- sapply(kegg.mm$kg.sets[nrlist$greater$allSets],length)
```

There are too many gene sets (`r dim(overlaps)[1]`), graph will be too crowded. Using non-redundant sets instead.  

```{r}
idx <- nrlist$greater$allSets %in% nrlist$greater$essentialSets
overlaps <- overlaps[idx,idx]
labels <- labels[idx]
sizes <- sizes[idx]
```

Now onto graphs.  

```{r fig.height=8, fig.width=10}
set.seed(50)
#g <- graph_from_adjacency_matrix(overlaps)
g <- as_tbl_graph(overlaps) %>% mutate(genes=sizes) %>%
  mutate(cluster = as.factor(group_infomap()))
ggraph(g, layout = "nicely") +
  geom_edge_link(alpha=0.2,edge_colour="blue", show.legend=FALSE) +
  geom_node_point(aes(size=genes,colour=cluster),alpha=0.9,show.legend=TRUE) +
  geom_node_text(aes(label=labels), repel=TRUE, size=3) + 
  theme_void() +
  scale_size(range=c(3,8))
```

### Extract and visualize the gene expression data for top gene sets  

The following code loops over the groups of resulting gene sets, creating a directory for each group in which the output files will be saved.  
Since `same.dir=FALSE` there is just one group, labeled `greater`, otherwise there would be both `greater` and `less`, containing upregulated and downregulated gene sets, respectively.  
The following set of operations is then executed for each non-redundant (i.e. essential) gene set:  
 * identify essential genes in the set
 * extract expression data (i.e. LFC values) for those genes
 * create output files (LFC and pathway maps)  

```{r message=FALSE, warning=FALSE, results='hide'}
require(pathview)
exprlist <- lapply(names(nrlist), function(x){
  
  # make output dir
  name <- x
  x <- nrlist[[name]]
  wdir <- getwd()
  outdir <- file.path("output",name)
  dir.create(outdir)
  setwd(outdir)
  
  # handle empty and matrix input
  if (!is.null(x)) {
    if (is.matrix(x)) {
      gene_sets <- rownames(x)
      x <- list(stats=x)
      x$coreGeneSets[[gene_sets]] <- kegg.mm.id[[gene_sets]]
    } else {
      gene_sets <- x$essentialSets
    }
    
    exprout <- lapply(gene_sets, function(gs){
      
      # gene set identifier
      gsname <- stringr::str_extract(gs,"mmu\\d+")
      
      # essential genes identifiers
      gsgenes <- x$coreGeneSets[[gs]]
      gsgenes <- gsgenes[!is.na(gsgenes)]
      
      # extract expression data for essential genes
      if (class(exp_data)=="numeric") 
        essData <- cbind(essGene(gsgenes, cbind(exp_data)))
      else
        essData <- essGene(gsgenes, exp_data)
      
      # output and visualize expression data
      gage::geneData(
        genes=kegg.mm.id[[gs]], exprs=essData, outname=gsname,
        txt=TRUE, heatmap=TRUE, scatterplot=TRUE
      )
      pathview(
        gene.data=exp_data, pathway.id=gs, species="mouse", gene.idtype="SYMBOL",
        kegg.native=TRUE, same.layer=FALSE,
        low=list(gene="#BC001F"), mid=list(gene="#E5E5E3"), high=list(gene="#0C74EB")
      )
      
      # expression data for ridge plot
      genes <- rownames(essData)
      ind <- which(genes %in% data$Name)
      expdata <- resdata[ind][,gene:=genes]
      if (nrow(expdata)>0) expdata else NULL
      
    })  
    
    names(exprout) <- gene_sets
  }
  
  setwd(wdir)
  exprout
  
})
```

Ridge plot for gene sets:  

```{r fig.height=10, fig.width=8}
allsets <- unlist(exprlist,recursive=FALSE)
dt <- rbindlist(allsets, idcol="geneset")
ggplot(dt, aes(x=log2FoldChange, y=geneset)) + 
  geom_density_ridges() +
  geom_vline(xintercept=0, color='firebrick', linetype='dashed') +
  xlab("LFC") + ylab(NULL) + DOSE::theme_dose()
```
